---
title: "Affymetrix Workflow"
author: "M. Kohl"
date: "April 2023"
output:
  rmarkdown::html_document:
    toc: true
    number_sections: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
## setup für die Chunks
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
```


# Introduction


# Import of Data

Download GSE2634 from Gene Expression Omnibus:
https://www.ncbi.nlm.nih.gov/geo/

Direct Link:
https://www.ncbi.nlm.nih.gov/geo/download/?acc=GSE2634&format=file

You might need some archive software to extract the data; see for instance
7zip (http://www.7-zip.de/)

Extract the files in a subfolder with name "Data". 
Remove all files except for GSM50690.CEL.gz, GSM50691.CEL.gz, ..., GSM50697.CEL.gz


```{r}
library(affy)
```


```{r, cache = TRUE}
## List CEL files
FileNames <- list.files("./Data/", full.names = TRUE)
## Import CEL files
affyData <- ReadAffy(filenames = FileNames)
affyData
```


# Quality Control

```{r, eval=FALSE}
library(arrayQualityMetrics)
arrayQualityMetrics(affyData, outdir = "QC", force = TRUE, 
                    do.logtransform = TRUE)
```


# Preprocessing

```{r}
library(gcrma)
library(vsn)
library(RobLoxBioC)
library(plier)
```


```{r, cache = TRUE}
affyData.rma <- rma(affyData)
affyData.gcrma <- gcrma(affyData)
affyData.vsnrma <- vsnrma(affyData)
affyData.mas5 <- robloxbioc(affyData, normalize = TRUE)
affyData.plier <- justPlier(affyData, normalize = TRUE)
```


# Quality Control of Preprocessed Data

```{r}
library(vsn)
meanSdPlot(affyData.rma)
meanSdPlot(affyData.gcrma)
meanSdPlot(affyData.vsnrma)
meanSdPlot(log2(exprs(affyData.mas5)))
meanSdPlot(affyData.plier)
```


```{r}
library(MKomics)
cor.mas5 <- cor(log2(exprs(affyData.mas5)))
cor.rma <- cor(exprs(affyData.rma))
cor.gcrma <- cor(exprs(affyData.gcrma))
cor.plier <- cor(exprs(affyData.plier))
cor.vsnrma <- cor(exprs(affyData.vsnrma))
```


```{r}
corPlot2(cor.mas5, minCor = 0.75, labels = pData(affyData)[,1], title = "MAS 5.0")
corPlot2(cor.rma, minCor = 0.85, labels = pData(affyData)[,1], title = "RMA")
corPlot2(cor.vsnrma, minCor = 0.80, labels = pData(affyData)[,1], title = "VSN-RMA")
corPlot2(cor.gcrma, minCor = 0.85, labels = pData(affyData)[,1], title = "GC-RMA")
corPlot2(cor.plier, minCor = 0.50, labels = pData(affyData)[,1], title = "PLIER")
```


```{r}
## range = 0: whiskers to most extreme data points
boxplot(log2(exprs(affyData.mas5)), range = 0, main = "MAS 5.0")
boxplot(exprs(affyData.rma), range = 0, main = "RMA")
boxplot(exprs(affyData.gcrma), range = 0, main = "GC-RMA")
boxplot(exprs(affyData.vsnrma), range = 0, main = "VSN-RMA")
boxplot(exprs(affyData.plier), range = 0, main = "PLIER")
```


# (Unspecific) Filtering

```{r}
library(genefilter)
SD.rma <- rowSds(exprs(affyData.rma))
SD.gcrma <- rowSds(exprs(affyData.gcrma))
SD.vsnrma <- rowSds(exprs(affyData.vsnrma))
```


```{r}
boxplot(list(RMA = SD.rma, GCRMA = SD.gcrma, VSNRMA = SD.vsnrma),
        range = 0)
```


```{r}
cut.rma <- quantile(SD.rma, prob = 0.75)
RMA.fi <- exprs(affyData.rma)[SD.rma > cut.rma,]
cut.gcrma <- quantile(SD.gcrma, prob = 0.75)
GCRMA.fi <- exprs(affyData.gcrma)[SD.gcrma > cut.gcrma,]
cut.vsnrma <- quantile(SD.vsnrma, prob = 0.75)
VSNRMA.fi <- exprs(affyData.vsnrma)[SD.vsnrma > cut.vsnrma,]
```


```{r}
meanSdPlot(RMA.fi)
meanSdPlot(GCRMA.fi)
meanSdPlot(VSNRMA.fi)
```


```{r}
cor.rma <- cor(RMA.fi)
cor.gcrma <- cor(GCRMA.fi)
cor.vsnrma <- cor(VSNRMA.fi)
corPlot2(cor.rma, minCor = 0.7, labels = pData(affyData)[,1], title = "RMA")
corPlot2(cor.gcrma, minCor = 0.78, labels = pData(affyData)[,1], title = "GC-RMA")
corPlot2(cor.vsnrma, minCor = 0.78, labels = pData(affyData)[,1], title = "VSN-RMA")
```


# Statistical Analysis

```{r}
group <- factor(c(rep("A1", 4), rep("A2", 4)))
```


```{r}
library(exactRankTests)
wfun <- function(x, g){ 
  wilcox.exact(x~g)$p.value
}
wilcox.pval <- apply(RMA.fi, 1, wfun, g = group)
```


```{r}
hist(wilcox.pval, breaks = seq(from = 0, to = 1, by = 0.01))
```


```{r}
library(multtest)
wilcox.pval.adj <- mt.rawp2adjp(wilcox.pval)
colSums(wilcox.pval.adj$adjp < 0.05)
```


```{r}
tfun <- function(x, g){
  t.test(x~g)$p.value
}
t.pval <- apply(RMA.fi, 1, tfun, g = group)
```


```{r}
hist(t.pval, breaks = seq(from = 0, to = 1, by = 0.01))
```


```{r}
t.pval.adj <- mt.rawp2adjp(t.pval)
colSums(t.pval.adj$adjp < 0.05)
```


```{r}
library(limma)
library(MKomics)
res <- mod.t.test(x = RMA.fi, group = group)
mod.t.pval.adj <- mt.rawp2adjp(res$p.value)
colSums(mod.t.pval.adj$adjp < 0.05)
```


```{r}
library(MKinfer)
volcano(res$`difference in means`, pval = res$adj.p.value, 
        effect0 = 0, effect.low = -1, effect.high = 1, 
        alpha = 0.25, xlab = "log2-fold Change (log2-FC)",
        ylab = expression(paste(-log[10], "(adj. p Value)")))
```


```{r}
library(ComplexHeatmap)
library(circlize)
library(RColorBrewer)
sel <- res$adj.p.value < 0.05 & abs(res$`difference in means`) > 1
selData <- RMA.fi[sel,]
selData <- selData - rowMeans(selData, na.rm = TRUE)
colnames(selData) <- as.character(group)
col1 <- rev(brewer.pal(n = 8, name = "RdYlBu"))
col2 <- brewer.pal(n = 3, name = "Set1")[c(3,1)]
mycol <- colorRamp2(seq(-2, 2, length = 128), colorRampPalette(col1)(128))
df <- data.frame(Group = group)
ha <- HeatmapAnnotation(df = df,
                        col = list(Group = c("A1" = col2[1],
                                             "A2" = col2[2])))
Heatmap(selData, col = mycol, name = "log2-FC", show_row_names = FALSE,
        top_annotation = ha, show_column_names = TRUE,
        column_names_gp = gpar(fontsize = 8),
        clustering_distance_columns = "pearson",
        show_row_dend = TRUE, 
        cluster_columns = TRUE, show_column_dend = TRUE,
        column_title = "Adj. p Value < 0.05 and |log2-FC| > 1",
        show_heatmap_legend = TRUE)
```


# Enrichment Analysis


```{r}
library(org.Hs.eg.db)
library(hgu133plus2.db)
GeneID <- select(hgu133plus2.db, keys = keys(hgu133plus2.db), 
                 columns = "ENTREZID")
keys.UP <- rownames(res)[(res$`difference in means` > 0 & 
                            res$adj.p.value < 0.05)]
keys.DN <- rownames(res)[(res$`difference in means` < 0 & 
                            res$adj.p.value < 0.05)]
GeneID.UP <- select(hgu133plus2.db, keys = keys.UP, columns = "ENTREZID")
GeneID.DN <- select(hgu133plus2.db, keys = keys.DN, columns = "ENTREZID")
```


```{r}
go <- goana(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
topGO(go, sort = "up")
topGO(go, sort = "down")
```


```{r}
kg <- kegga(list(Up = unique(GeneID.UP$ENTREZID), 
                 Down = unique(GeneID.DN$ENTREZID)), 
            universe = unique(GeneID$ENTREZID), 
            species = "Hs")
topKEGG(kg, sort = "up")
topKEGG(kg, sort = "down")
```


```{r}
library(simplifyEnrichment)
library(GO.db)
library(org.Hs.eg.db)
egGO2ALLEGS <- getFromNamespace("org.Hs.egGO2ALLEGS", "org.Hs.eg.db")
entrez_list <- unique(GeneID$ENTREZID)
entrez_list <- entrez_list[order(as.numeric(entrez_list))]
entrez_list <- entrez_list[!is.na(entrez_list)]

AnnotationDbi::Lkeys(egGO2ALLEGS) <- entrez_list
GeneID.PathID <- AnnotationDbi::toTable(egGO2ALLEGS)[, c("gene_id", "go_id", "Ontology")]
d <- duplicated(GeneID.PathID[, c("gene_id", "go_id")])
GeneID.PathID <- GeneID.PathID[!d, ]

entrez.sig.up <- unique(GeneID.UP$ENTREZID)
entrez.sig.up <- entrez.sig.up[order(as.numeric(entrez.sig.up))]
entrez.sig.up <- entrez.sig.up[!is.na(entrez.sig.up)]
GO.sig.up <- GeneID.PathID[GeneID.PathID$gene_id %in% entrez.sig.up,]

entrez.sig.dn <- unique(GeneID.DN$ENTREZID)
entrez.sig.dn <- entrez.sig.dn[order(as.numeric(entrez.sig.dn))]
entrez.sig.dn <- entrez.sig.dn[!is.na(entrez.sig.dn)]
GO.sig.dn <- GeneID.PathID[GeneID.PathID$gene_id %in% entrez.sig.dn,]

## very long computations
#sim.BP.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "BP"], ont = "BP")
#sim.CC.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "CC"], ont = "CC")
#sim.MF.up <- GO_similarity(GO.sig.up$go_id[GO.sig.up$Ontology == "MF"], ont = "MF")
#sim.BP.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "BP"], ont = "BP")
#sim.CC.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "CC"], ont = "CC")
sim.MF.dn <- GO_similarity(GO.sig.dn$go_id[GO.sig.dn$Ontology == "MF"], ont = "MF")

## very long computations
#res.BP.up <- simplifyGO(sim.BP.up, column_title = "GO BP - up")
#res.CC.up <- simplifyGO(sim.CC.up, column_title = "GO CC - up")
#res.MF.up <- simplifyGO(sim.MF.up, column_title = "GO MF - up")
#res.BP.dn <- simplifyGO(sim.BP.dn, column_title = "GO BP - down")
#res.CC.dn <- simplifyGO(sim.CC.dn, column_title = "GO CC - down")
res.MF.dn <- simplifyGO(sim.MF.dn, column_title = "GO MF - down")
```


```{r}
library(clusterProfiler)
ego <- enrichGO(gene          = GO.sig.dn$gene_id,
                universe      = entrez_list,
                OrgDb         = org.Hs.eg.db,
                ont           = "MF",
                pAdjustMethod = "BH",
                pvalueCutoff  = 0.2,
                qvalueCutoff  = 0.2,
                readable      = TRUE)
head(ego)
goplot(ego)
```


```{r}
library(DOSE)
edo <- enrichDGN(GO.sig.up$gene_id, 
                 universe = entrez_list,
                 pvalueCutoff = 0.2)
library(enrichplot)
barplot(edo)
dotplot(edo)
edox <- setReadable(edo, 'org.Hs.eg.db', 'ENTREZID')
p1 <- cnetplot(edox, foldChange=GO.sig.up$gene_id)
p1
p2 <- cnetplot(edox, foldChange=GO.sig.up$gene_id, 
               circular = TRUE, colorEdge = TRUE)
p2
p3 <- cnetplot(edox, node_label="category", cex_label_category = 1.2)
p3
p4 <- cnetplot(edox, node_label="gene", cex_label_gene = 0.8)
p4
p5 <- heatplot(edox)
p5
edox2 <- pairwise_termsim(edox)
p6 <- treeplot(edox2)
p6
```


# Software

```{r}
sessionInfo()
```
